Palomar 5 Gaps

### 2024-11-07 paper/main.tex
edited methods

  \subsection{Numerical Methodology}
    Our numerical methodology follows the approach outlined in Ferrone et al. (2023), which we present here for completeness. We begin by extracting the phase space properties (positions, velocities) as well as the masses and half-mass radii of 165 globular clusters from the galactic globular cluster catalog by \cite{2021MNRAS.505.5957B}.\footnote{The Baumgardt catalog has been assembled across a series of works, see: \cite{2020PASA...37...46B,2019MNRAS.482.5138B,2018MNRAS.478.1520B} as well. The catalog can be found on the World Wide Web at \href{https://people.smp.uq.edu.au/HolgerBaumgardt/globular/}{https://people.smp.uq.edu.au/HolgerBaumgardt/globular/}.} We handle the uncertainties on the measured phase space parameters through a monte-carlo approach. The catalog provides uncertainties for mass, distances, line of sight velocities, proper motions and includes a covariance term between the proper motions---uncertainties for the right ascension, declination, and the characterisitcs radii are not available. We repeate simulation the creation of Palomar 5's tidail tails 50 times, for each of which we vary the initial conditions of each cluster by sampling its five-dimensional gaussian uncertaintiy distribution. For the first simulation, which we name \texttt{Sampling 000}, we use the most probable values for the initial conditions.

    We convert the initial conditions from sky coordinates into a galactocentric reference frame. We use the local standard of rest and the peculiary motion of the sun as reported by Sch\"onrich 2010. The circular velocity at the solar radius was set to: $v_{\text{LSR}} = 240$~km~s$^{-1}$. The peculiar velocity of the sun was set to $(U_\odot, V_\odot, W_\odot)=(11.1, 12.24, 7.25)$~km~s$^{-1}$. The position of the Sun was set to $(x_\odot,y_\odot,z_\odot) = (-8.34,0,0.027)$~kpc, vertical position above the disk was taken from Chen et al 2001 and the galacto-radial distance was taken from Reid et al 2014. These transformations were performed using \texttt{astropy} \citep{2013A&A...558A..33A}.

    For the galactic potential, we used the second static model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table 1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. \salvatore{This model satisfies a series of observational constraints such as local solar stellar density, the galactic rotation curve, etc, similarly to other galactic models such as Bovy2014, McMillian.} However, To balance data volume and computation time, we use only one galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models, with kinematic uncertainties being the dominant factor in orbital solution differences. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be model-independent, similarly to \citet{2023A&A...673A..44F}. While the clusters' positions may depend on the model, we assert that interaction rates and stream formation are unaffected by the choice of galactic model.

    There are two primary methodological departures from \citet{2023A&A...673A..44F}. Previously, the clusters only experienced the galaxy. Now, we include N-body forces between the clusters. Secondly, the clusters are not point masses but are softened as Plummer spheres each with their own scale radii. The equations of motion for the globular cluster system are thus: 
    \begin{equation}
      \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
    \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galacto-centric position, the index $i$ indicates a globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates. The index $j$ indicates the other globular cluster that are summed over where: $m_j$ is its mass, $b_j$ is its Plummer scale radius, and $r_j$ is it's galacto-centric position. $\Phi$ is the axis-symmteric potential of the model used for the Milky Way. Note that the mass and scale radii are constant in these simulations and we do not take into account the internal evolution of the globular clusters.

    At the end of this integration we have positions and velocities of the entire globular cluster system 5 Gyr ago. We sample 100,000 particles from a Plummer distribution using the same total mass and characteristic radius for the orbital integration. We repeate this sampling for each of the 50 Monte-Carlo initial conditions. Considering the mass uncertainties is an improvement from \citet{2023A&A...673A..44F} where the most probable mass was used for each Monte-Carlo sampling.

    Regarding Palomar 5's mass, we use a mass estimate for Palomar 5 taken from \citet{2021MNRAS.505.5957B}, $1.34\pm0.24\times10^{4}M_{\odot}$--which is a contested value.

    We then integrate the evolution of these particles forward in time to the present day. During this forward integration, each star particle feels the gravitational influence of the galaxy and all globular clusters, but not other star particles. We do not perform direct N-body simulations between star particles as we did for globular clusters. In other words, we are using test-particle and solving the restricted three body problem. The particles do not exert/feel gravitational forces upon/from one another. At each time step, we load the positions of all globular clusters, as found from the initial integration, and apply their forces to the star particles. The equation of motion is thus: 
    \begin{equation}
      \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
      \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}

    The integration was performed using a leapfrog integrator. The timestep size was chosen to be based on two critera. First, to ensure the conservation of energy to XXX. The second, is the drift in position due to numerical error over 5 Gyr was negligible compared to the scale radius of the globular clusters, which is about 5~pc. Further details on the timestep selection are discussed in the appendix.

@@

@@

  \salvatore{I'm a bit lost on trying to reconcile the differences between Baumgardt's estiamte and \citet{2017ApJ...842..120I}'s estimate in sections 5.2 and 5.3. It looks like 5.2 is based on counts and then supplementing the completness function with dNdM \citet{2001AJ....122.3231G} and demonstrating that Salpeter would overestimate the mass. I think 5.3 is showing that an MCMC was used to add more context to the estimate from 5.3.}

>>>>

  \subsection{Numerical Methodology}
    Our numerical methodology follows the approach outlined in \citet{2023A&A...673A..44F}, which we present here for completeness. We begin by extracting the phase space properties (positions, velocities) as well as the masses and half-mass radii of 165 globular clusters from the galactic globular cluster catalog by \cite{2021MNRAS.505.5957B}.\footnote{The Baumgardt catalog has been assembled across a series of works, see: \cite{2020PASA...37...46B,2019MNRAS.482.5138B,2018MNRAS.478.1520B} as well. The catalog can be found on the World Wide Web at \href{https://people.smp.uq.edu.au/HolgerBaumgardt/globular/}{https://people.smp.uq.edu.au/HolgerBaumgardt/globular/}.} We handle the uncertainties on the measured phase space parameters through a monte-carlo approach. The catalog provides uncertainties for mass, distances, line of sight velocities, proper motions and includes a covariance term between the proper motions---uncertainties for the right ascension and declination are negligible and the characterisitcs radii are not available. We sample the these five-dimensional gaussians for each cluster 50 times. For the first simulation, which we name \texttt{Sampling 000}, we use the most probable values for the initial conditions.

    We convert the initial conditions from sky coordinates into a galactocentric reference frame. We use the local standard of rest and the peculiary motion of the sun as reported by Sch\"onrich 2010. The circular velocity at the solar radius was set to: $v_{\text{LSR}} = 240$~km~s$^{-1}$. The peculiar velocity of the sun was set to $(U_\odot, V_\odot, W_\odot)=(11.1, 12.24, 7.25)$~km~s$^{-1}$. The position of the Sun was set to $(x_\odot,y_\odot,z_\odot) = (-8.34,0,0.027)$~kpc, vertical position above the disk was taken from \citet{2001ApJ...553..184C} and the galacto-radial distance was taken from \citet{2004ApJ...616..872R}. These transformations were performed using \texttt{astropy} \citep{2013A&A...558A..33A}.

    For the galactic potential, we used the second static model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table~1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. This model satisfies a series of observational constraints such as local solar stellar density, the galactic rotation curve, etc, similarly to other galactic models such as Bovy2014, McMillian. However, To balance data volume and computation time, we use only one galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models, with kinematic uncertainties being the dominant factor in orbital solution differences. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be model-independent, similarly to \citet{2023A&A...673A..44F}. While the clusters' positions may depend on the model, we assert that interaction rates and stream formation are unaffected by the choice of galactic model.

    The primary methodological departure from \citet{2023A&A...673A..44F} is that in the first integration step, when we integrate the globular clusters backward by five billion years, we compute N-body forces between them. Specifically, the clusters are not point masses but are softened as Plummer spheres each with their own scale radii. The equations of motion for the globular cluster system are thus: 
    \begin{equation}
      \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
    \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galacto-centric position, the index $i$ indicates a globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates. The index $j$ indicates the other globular cluster that are summed over where: $m_j$ is its mass, $b_j$ is its Plummer scale radius, and $r_j$ is it's galacto-centric position. $\Phi$ is the axis-symmteric potential of the model used for the Milky Way. Note that the mass and scale radii are constant in these simulations and we do not take into account the internal evolution of the globular clusters.

    At the end of this integration we have positions and velocities of the entire globular cluster system 5 Gyr ago. We sample 100,000 particles from a Plummer distribution using the same total mass and characteristic radius for the orbital integration. We repeate this sampling for each of the 50 Monte-Carlo initial conditions. Considering the mass uncertainties is an improvement from \citet{2023A&A...673A..44F} where only the most probable mass was used for each Monte-Carlo sampling.

    This marks the begining of the second step of our procedure where we then integrate the evolution of these particles forward in time to the present day. During this forward integration, each star particle feels the gravitational influence of the galaxy and all globular clusters, but not other star particles. We do not perform direct N-body simulations between star particles as we did for globular clusters. In other words, we are using test-particle and solving the restricted three body problem. The particles do not exert/feel gravitational forces upon/from one another. At each time step, we load the positions of all globular clusters, as found from the initial integration, and apply their forces to the star particles. The equation of motion is thus: 
    \begin{equation}
      \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
      \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}

    The integration was conducted using a leapfrog integrator, chosen for its ability to preserve phase-space volume and conserve the Hamiltonian with each integration step. This method is preferable to, for instance, a Runge-Kutta scheme, which can introduce non-physical and significant numerical errors in systems that require long-term stability and energy conservation. One drawback of the leapfrog integrator is that it requires a uniform timestep across the entire system, leading to unnecessary computations for slower-moving particles compared to adaptive timestep methods. However, energy conservation and phase-space volume preservation are paramount when modeling stellar streams. The timestep was therefore set small enough to ensure energy conservation for all particles within the host globular cluster, a value that depends on the system and scales inversely with its concentration.

@@

  This procedure was conducted on Palomar 5 with 50 different samplings of the initial conditions, producing 50 distinct realizations of the cluster's orbits. For these simulations, a timestep of 10,000 years was selected to maintain adequate energy conservation and enable \textit{reverse integrability} of the globular cluster system. By reverse integrability, we mean the integrator's capability to track the cluster backward in time and then re-integrate it forward along the same trajectory. The drift in the final position after forward integration, compared to the initial position from the backward integration, was consistently at least two orders of magnitude smaller than the Plummer scale radius used for Palomar 5. This high precision ensures that no fictitious evil numerical forces influenced the system, preventing any artificial mass loss or retention of star particles.

@@

  \salvatore{I'm a bit lost on trying to reconcile the differences between Baumgardt's estiamte and \citet{2017ApJ...842..120I}'s estimate in sections 5.2 and 5.3. It looks like 5.2 is based on counts and then supplementing the completness function with dNdM \citet{2001AJ....122.3231G} and demonstrating that Salpeter would overestimate the mass. I think 5.3 is showing that an MCMC was used to add more context to the estimate from 5.3.}  Regarding Palomar 5's mass, we use a mass estimate for Palomar 5 taken from \citet{2021MNRAS.505.5957B}, $1.34\pm0.24\times10^{4}M_{\odot}$--which is a contested value.
